---
title: "Spatial transcriptomics analysis of FFPE pancreas of human"
author: "Frano Malinarich G"
Company: "Doppelganger Biosystem GmbH"
date: "`r Sys.Date()`"
format:
  html:
    toc: true      
    toc-depth: 4   
    number-sections: true  
    fig-width: 8    
    fig-height: 6   
    code-fold: true 
    code-download: true 
    css: 
      - styles.css       
editor: visual
---

# **Spatial transcriptomics analysis of human FFPE pancreas sample**

### Author of the analysis: *Frano Malinarich G. Ph.D.*

## Overall description

<https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-human-pancreas-ffpe-dataset/>

"With CosMx™ Spatial Molecular Imager (SMI) we characterized human pancreas FFPE tissue with a 18,946 plex pre-commercial version of our CosMx Human Whole Transcriptome panel. 
This pre-commercial whole transcriptome panel showcases the highest plex of any spatial imager. 
With over 18,946 genes, this RNA panel offers researchers the ability to analyze the entire transcriptome at true single cell level at the highest plex." 

## Specific description (See Table)

| Parameter                                         | Value                                                  |
|----------------------------------|--------------------------------------|
| Tissue Type                                       | Pancreas                                               |
| Panel                                             | Pre-commercial WTx RNA panel containing 18,946 targets |
| Scan Area                                         | 4.695 mm²                                              |
| Total Cells                                       | 48951                                                  |
| \% Cells Passing QC                               | 96.6%                                                  |
| Total transcripts detected                        | 64960677                                               |
| \% of transcripts assigned to a cell              | 91%                                                    |
| Sensitivity (Mean transcripts/cell)               | 1214                                                   |
| Sensitivity (Mean transcripts per µm3)            | 3.06                                                   |
| Genomic breadth (Mean unique genes per cell)      | 479                                                    |
| Genomic breadth (Unique genes detected above LOD) | 6710                                                   |
| Mean negatives / Plex / cell                      | 0,011                                                  |
| Mean SystemControl / Plex / cell                  | 0,005                                                  |

## Platforms:

-   CosMx SMI (Spatial Molecular Imager) Technology

## Aims

0.  Connect AWS-EC2 instance to local computer.
1.  Import CosMx dataset into AWS-EC2 and and decompress it
2.  Load the CosMx dataset into RStudio Server (continue by creating seurat object, preprocessing, etc).
3.  Visualizations: UMAP and tissue/cell type characterization
4.  Tabular visualization of exportable file
5.  Convert seurat object metadata in H5AD

## *Activate packages*

```{r}
# Load/install packages

if (!requireNamespace("pacman")) {
  install.packages("pacman")
}
pacman::p_load(conflicted, dplyr, tibble, Seurat,
               patchwork, ggplot2, ggrepel, stringr, 
               gridExtra, tidyr, flextable,
               officer, rmarkdown, R.utils,
               here, webshot2, harmony, future
               )


# Resolve conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("combine", "dplyr")
conflict_prefer("setdiff", "base")  # Specify base::setdiff to avoid errors
conflict_prefer("rename", "dplyr")
conflict_prefer("width", "flextable")  # For table formatting
conflict_prefer("select", "dplyr")
conflict_prefer("arrange", "dplyr")
conflict_prefer("count", "dplyr")
conflict_prefer("slice", "dplyr")
conflict_prefer("Assays", "Seurat")

library(cowplot) # For plot arrangement
library(SpatialExperiment)
library(Matrix)
library(SeuratWrappers)
library(reticulate)
library(future)
library(pagedown)
library(sf)   # for handling polygons
library(data.table)
library(glmGamPoi)
packageVersion("glmGamPoi") # Should be 1.12.0 or higher

```

## *Load spatial data and QC*

```{r}

# Steps

# Load Nanostring CosMx dataset (Pancreas) and save seurat object
nano.obj <- LoadNanostring(
  data.dir = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/RData",
  fov = "Pancreas",
  assay = "Nanostring"
)

nano.obj

# An object of class Seurat 
# 21731 features across 48944 samples within 1 assay 
# Active assay: Nanostring (21731 features, 0 variable features)
#  1 layer present: counts
#  1 spatial field of view present: Pancreas

# Save object
# saveRDS(
#   nano.obj,
#   file = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_pancreas.rds")


# Checking whether the seurat object has Spatial coordinates incorporated
names(nano.obj@images)

slotNames(nano.obj@images[["Pancreas"]])

head(nano.obj@images[["Pancreas"]]@boundaries$centroids)
head(nano.obj@images[["Pancreas"]]@boundaries$centroids@cells)
# head(nano.obj@images[["Pancreas"]]@boundaries$centroids@nsides)
# head(nano.obj@images[["Pancreas"]]@boundaries$centroids@radius)
# head(nano.obj@images[["Pancreas"]]@boundaries$centroids@theta)
head(nano.obj@images[["Pancreas"]]@boundaries$centroids@coords)
# head(nano.obj@images[["Pancreas"]]@boundaries$centroids@bbox)
# head(nano.obj@images[["Pancreas"]]@boundaries$centroids@proj4string)

head(nano.obj@images[["Pancreas"]]@boundaries$segmentation)
head(nano.obj@images[["Pancreas"]]@boundaries$segmentation@polygons)
# head(nano.obj@images[["Pancreas"]]@boundaries$segmentation@plotOrder)
# head(nano.obj@images[["Pancreas"]]@boundaries$segmentation@bbox)
# head(nano.obj@images[["Pancreas"]]@boundaries$segmentation@proj4string)

# head(GetTissueCoordinates(nano.obj), 3) # THIS IS FOR 10X VISIUM!!!

# str(nano.obj@images[["F058"]]@coordinates) # WRONG! It has to be "Pancreas"

view(nano.obj@meta.data)

# Conclusion: the seurat object, according to slot "boundaries" has all the spatial coordinates.

# Read seurat object
# nano.obj <- readRDS("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_pancreas.rds")

# ---------------------------------- QC ---------------------------------- 

# # Add mitochondrial % (CosMx may use human symbols, e.g., MT-CO1, MT-ND1)
nano.obj[["percent.mt"]] <- PercentageFeatureSet(nano.obj, pattern = "^MT-")

# # Add ribosomal % (for human symbols: RPL*, RPS*)
nano.obj[["percent.rb"]] <- PercentageFeatureSet(nano.obj, pattern = "^RP[SL]")

head(nano.obj@meta.data, 4)

# Violin plot QC
VlnPlot(nano.obj, features = c("nCount_Nanostring", "nFeature_Nanostring",
                               "percent.mt", "percent.rb"), pt.size = 0.1) + NoLegend()

# Scatter QC
FeatureScatter(nano.obj, feature1 = "percent.rb", feature2 = "nCount_Nanostring") +
FeatureScatter(nano.obj, feature1 = "percent.mt", feature2 = "nCount_Nanostring")

# Summary statistics
summary(nano.obj@meta.data[, c("nFeature_Nanostring", "nCount_Nanostring", "percent.mt", "percent.rb")])


# Save object
# saveRDS(
#   nano.obj,
#   file = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_QC.rds")


```

## *Data pre-processing: Data cleaning & Preprocessing: SCTransfomation - FindVariableFeatures*

```{r}

# Read seurat object
# nano.obj <- readRDS("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_QC.rds")

nano.obj
# An object of class Seurat
# 21731 features across 48944 samples within 1 assay
# Active assay: Nanostring (21731 features, 0 variable features)
#  1 layer present: counts
#  1 spatial field of view present: Pancreas

# Assays(nano.obj) # "Nanostring"

# ---------------------------- Cleaning: identifying/removal of synthetic genes ----------------------------

# The purpose of cleaning is to identify and remove all synthetic genes, which are often artificial spike-ins or placeholder features from imaging pipelines (e.g., CosMx/NanoString) and shouldn't be considered in biological interpretation or downstream analysis. 

possible_patterns <- c("SystemControl", "NegPrb", "PosPrb", "Blank", "Housekeeping", "ERCC")

for (pattern in possible_patterns) {
  hits <- grep(pattern, rownames(nano.obj), ignore.case = TRUE, value = TRUE)
  if (length(hits) > 0) {
    cat(sprintf("Found %d genes matching '%s'\n", length(hits), pattern))
    print(head(hits, 5))
  } else {
    cat(sprintf("No genes matching '%s'\n", pattern))
  }
}

# Identify SystemControl genes only (synthetic controls to remove)
control_genes <- grep("^SystemControl", rownames(nano.obj), ignore.case = TRUE, value = TRUE)

# Remove from all assays
nano.obj_clean <- nano.obj
for(assay_name in Assays(nano.obj)) {
  # Get current features
  all_features <- rownames(GetAssayData(nano.obj, assay = assay_name))

  # Filter out control genes
  keep_features <- setdiff(all_features, control_genes)

  # Subset the assay
  nano.obj_clean[[assay_name]] <- subset(nano.obj[[assay_name]], features = keep_features)
}

# Verify removal
cat("Original gene count:", nrow(GetAssayData(nano.obj, assay = "Nanostring")), "\n")
cat("Cleaned gene count:", nrow(GetAssayData(nano.obj_clean, assay = "Nanostring")), "\n")
cat("Control genes still present:", length(base::intersect(rownames(GetAssayData(nano.obj_clean, assay = "Nanostring")), control_genes)), "\n")


nano.obj_clean
# An object of class Seurat 
# 18996 features across 48944 samples within 1 assay 
# Active assay: Nanostring (18996 features, 0 variable features)
#  1 layer present: counts
#  1 spatial field of view present: Pancreas

# Save
# saveRDS(
#   nano.obj_clean,
#   file = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_clean.rds")


# ---------------------------- Preprocessing ----------------------------

# Read seurat object
# nano.obj <- readRDS("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_clean.rds")

nano.obj
# An object of class Seurat 
# 18996 features across 48944 samples within 1 assay 
# Active assay: Nanostring (18996 features, 0 variable features)
#  1 layer present: counts
#  1 spatial field of view present: Pancreas

# head(nano.obj@meta.data, 4)


# SCTransform
nano.obj <- SCTransform(
  nano.obj,
  assay = "Nanostring",
  clip.range = c(-10, 10),
  conserve.memory = TRUE,       # ✅ Definitely use (saves ~40% RAM)
  method = "glmGamPoi",         # ✅ Strongly recommended (faster + more accurate)
  variable.features.n = 5000,   # Increase from default 3000
  return.only.var.genes = FALSE,
  vars.to.regress = "percent.rb",
  verbose = TRUE,
)

nano.obj
# An object of class Seurat 
# 37992 features across 48944 samples within 2 assays 
# Active assay: SCT (18996 features, 5000 variable features)
#  3 layers present: counts, data, scale.data
#  1 other assay present: Nanostring
#  1 spatial field of view present: Pancreas

Assays(nano.obj) # "Nanostring" "SCT"

view(nano.obj@meta.data)

# colnames(nano.obj@meta.data) # [1] "orig.ident"  "nCount_Nanostring"  "nFeature_Nanostring" "nCount_SCT"         [5] "nFeature_SCT"
# head(row.names(nano.obj), 20)
# # [1] "A1BG"    "A1CF"    "A2M"     "A2ML1"   "A3GALT2" "A4GALT"  "A4GNT"   "AAAS"    "AACS"    "AADAC"
# # [11] "AADACL2" "AADACL3" "AADACL4" "AADAT"   "AAGAB"   "AAK1"    "AAMDC"   "AAMP"    "AANAT"   "AAR2"
# length(row.names(nano.obj)) # 18996



# -------------- Verifications ---------------

# Original genes in Nanostring assay
original_genes <- nrow(GetAssayData(nano.obj, assay = "Nanostring", layer = "counts")) # 18996

# Genes in SCT assay
sct_genes <- nrow(nano.obj[["SCT"]]) # 18996

cat(sprintf(
  "Genes retained: %.1f%% (%d/%d)",
  sct_genes/original_genes * 100,
  sct_genes,
  original_genes
))

# Expected Output:
# Genes retained: 100.0% (18996/18996)
# This confirms no genes were lost during SCT

# Top 10 variable genes
# head(VariableFeatures(nano.obj), 100)
# Expected: A list of biologically relevant pancreatic genes (e.g., INS, GCG, SST, TTR, PPY)

# Check residual distribution
feature.attrs <- SCTResults(nano.obj, slot = "feature.attributes")
summary(feature.attrs$residual_variance)
# Healthy distribution:
# Median (0.65355) within optimal 0.5-2 range
# Max variance (529.89) expected for highly variable genes like INS
# No zero-variance genes (Min > 0)

# Visualize Transformation Results
VlnPlot(nano.obj,
        features = c("nCount_Nanostring", "nCount_SCT"),
        group.by = "orig.ident",
        pt.size = 0)
# Expected: to see nCount_SCT with the sioullioutte like a human (with a waist)


# Save object
# saveRDS(
#   nano.obj,
#   file = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_pancreas_SCT.rds")


#############################################
###### IMPORTANT 
#############################################
# Skip "FindSpatiallyVariableFeatures" because is not necessary for our analysis. Also, it is adapted to 10x Visium.
# Skip this "RunPCA" with 'features = SpatiallyVariableFeatures()' has to be "VariableFeatures". 

# Identify variable features (optional but useful)
# nano.obj <- FindVariableFeatures(
#   nano.obj,
#   assay = "SCT",
#   nfeatures = 3000
#   )

# SCTransform() already selected 5000 variable features by default (variable.features.n = 5000).
# When you call FindVariableFeatures() again afterward, you're replacing those with a new set — in your case, selecting the top 3000. By reducing to 3000 can: Speed up computation, Slightly denoise the data,
# But may drop informative genes from spatially or biologically relevant contexts (especially in spatial data).
# Recommendation: Keep the SCTransform defaults (5000)
###########################################
###########################################

# nano.obj
# An object of class Seurat 
# 37992 features across 48944 samples within 2 assays 
# Active assay: SCT (18996 features, 5000 variable features)
#  3 layers present: counts, data, scale.data
#  1 other assay present: Nanostring
#  1 spatial field of view present: Pancreas


# colnames(nano.obj@meta.data)
# view(nano.obj@meta.data)
# head((nano.obj@meta.data), 3)
# colnames(nano.obj@images$Pancreas@boundaries$segmentation@bbox)

```

## *Dimensional reduction, clustering and visualizations*

```{r}
# Load SCTransformed Seurat object
# nano.obj <- readRDS("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_pancreas_SCT.rds")

# nano.obj
# An object of class Seurat 
# 37992 features across 48944 samples within 2 assays 
# Active assay: SCT (18996 features, 5000 variable features)
#  3 layers present: counts, data, scale.data
#  1 other assay present: Nanostring
#  1 spatial field of view present: Pancreas


# PCA on variable genes (not spatially variable)
nano.obj <- RunPCA(
  nano.obj,
  assay = "SCT",
  features = VariableFeatures(nano.obj)
  )

# 5️⃣ Visualize variance explained
ElbowPlot(nano.obj, ndims = 50)

# 6️⃣ Run UMAP, neighbors, and clustering
nano.obj <- RunUMAP(nano.obj, dims = 1:21, reduction = "pca")
nano.obj <- FindNeighbors(nano.obj, dims = 1:21)
nano.obj <- FindClusters(nano.obj, resolution = 1.0)

# 7️⃣ UMAP & spatial visualization of clusters
DimPlot(nano.obj, raster = FALSE, label = TRUE, pt.size = 1.2) + ggtitle("UMAP: Cell Clusters")
ImageDimPlot(nano.obj, cols = "glasbey") + ggtitle("Spatial Tissue Clusters")

nano.obj
# An object of class Seurat 
# 37992 features across 48944 samples within 2 assays 
# Active assay: SCT (18996 features, 5000 variable features)
#  3 layers present: counts, data, scale.data
#  1 other assay present: Nanostring
#  2 dimensional reductions calculated: pca, umap
#  1 spatial field of view present: Pancreas


# Save object
# saveRDS(
#   nano.obj,
#   file = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_pancreas_PCA-UMAP.rds")


```

## *Differential Expression Gene (DEG) analysis - cell annotation - visualizations & tables*

```{r}
# Load Seurat object
nano.obj <- readRDS("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_pancreas_PCA-UMAP.rds")
 
# nano.obj
# An object of class Seurat 
# 37992 features across 48944 samples within 2 assays 
# Active assay: SCT (18996 features, 5000 variable features)
#  3 layers present: counts, data, scale.data
#  1 other assay present: Nanostring
#  2 dimensional reductions calculated: pca, umap
#  1 spatial field of view present: Pancreas

# 8️⃣ Marker discovery
All.markers <- FindAllMarkers(
  nano.obj,
  assay = "SCT",
  only.pos = TRUE,            # Report only upregulated markers
  min.pct = 0.1,              # More inclusive (detects rare cell-type markers). 
  logfc.threshold = 0.1,      # Lower LFC threshold (more inclusive)
  test.use = "wilcox",        # Standard test for DE
  return.thresh = 0.01        # FDR < 1%
)

# min.pct = 0.1 (Default): This means the gene must be expressed in at least 10% of cells in either of the two clusters being compared.
# return.thresh = 0.01: Filters for adjusted p-value (FDR). Default is 0.05. Using 0.01 makes the result more stringent.


write.csv(All.markers, 
          "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/markers_tissue_clusters.csv", 
          row.names = FALSE
          )

# View top 5 markers per cluster
top_markers <- All.markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5)

# Display the result
head(top_markers, 10)

dim(All.markers)
head(All.markers)

# A tibble: 6 × 7
# Groups:   cluster [2]
#       p_val avg_log2FC pct.1 pct.2 p_val_adj cluster gene    
#       <dbl>      <dbl> <dbl> <dbl>     <dbl> <fct>   <chr>   
# 1 0               2.36 0.421 0.099 0         0       RBPJL   
# 2 0               2.27 0.349 0.081 0         0       PGGHG   
# 3 0               2.27 0.441 0.112 0         0       PNLIPRP2
# 4 0               2.20 0.168 0.042 0         0       FUT1    
# 5 1.31e-276       2.13 0.131 0.033 2.50e-272 0       IZUMO1  
# 6 0               3.28 0.296 0.043 0         1       MUC6   

# The table shows the high DEG from cells in cluster zero (then 1...) vs all the other clusters
# p_val: Raw p-value from the statistical test (Wilcoxon test here) comparing expression levels
# avg_log2FC: Log₂ fold-change between cluster vs. all other cells — higher = more specific marker
# pct.1:	% of cells in the cluster expressing the gene
# pct.2: % of cells in other clusters expressing the gene
# p_val_adj: 	Adjusted p-value (FDR corrected using Bonferroni or BH)
# cluster:	Cluster ID (as assigned by FindClusters)

# Recommendations:
# Don't rely only on log2FC. Look at both pct.1 vs. pct.2 to judge specificity.
# A gene with high avg_log2FC but expressed in very few cells (e.g., <10%) may not be biologically useful.
# Always validate marker genes against known biological expectations — like you’re doing with insulin, glucagon, etc.

dplyr::arrange(All.markers, desc(avg_log2FC)) %>% 
  dplyr::slice_head(n = 10)

#       p_val avg_log2FC pct.1 pct.2 p_val_adj cluster  gene
# GCG       0   9.426866 0.974 0.029         0      15   GCG
# TTR       0   9.150177 0.954 0.015         0      15   TTR
# INS1      0   6.772591 0.997 0.058         0      16   INS
# CHGA      0   6.515519 0.590 0.013         0      15  CHGA
# OLFM4     0   6.373070 1.000 0.042         0      13 OLFM4
# CHGB      0   6.118834 0.389 0.009         0      15  CHGB
# PLVAP     0   6.060987 0.791 0.028         0      10 PLVAP
# VGF       0   6.029267 0.371 0.010         0      15   VGF
# H4C6      0   5.867569 0.338 0.009         0      18  H4C6
# H1-5      0   5.864857 0.458 0.013         0      18  H1-5



# 9️⃣ Explore known cell-type markers

# # Endocrine
# FeaturePlot(nano.obj, features = c("INS", "GCG", "SST", "PPY", "GHRL"), label = TRUE)
# 
# # Exocrine
# FeaturePlot(nano.obj, features = c("CPA1", "PRSS1"), label = TRUE, pt.size = 1)  # Acinar
# FeaturePlot(nano.obj, features = c("KRT19", "CFTR"), label = TRUE, pt.size = 1)  # Ductal
# 
# # Immune
# FeaturePlot(nano.obj, features = c("LYZ", "CD3E", "CD14", "CD19"), 
#             label = TRUE, pt.size = 1)

# Violin plot
VlnPlot(nano.obj,
        features = c("INS", "GCG", "SST", "CPA1", "KRT19", "CD14", "CD3G" , "PECAM1"),
        assay = "Nanostring",
        layer = "counts",
        pt.size = 0.1) + NoLegend()

# 🔟 Curated Pancreatic Cell-Type Markers

panc_markers <- list(
  beta        = c("INS", "IAPP", "PDX1", "MAFA", "SLC30A8", "CHGA"),
  alpha       = c("GCG", "TTR", "IRX2", "FXYD2", "CRYBA2"),
  gamma       = c("GHRL", "SST", "LEP", "NPY", "PYY"),
  delta       = c("SST", "HHEX", "RBP4", "SEC11C", "LEPR"),
  PP          = c("PPY", "AQP3", "MEIS2", "PLA2G1B", "NR5A2"),
  acinar      = c("PRSS1", "CPA1", "CPA2", "CTRB2", "AMY2A", "MUC6"),
  ductal      = c("KRT19", "KRT8", "KRT7", "CFTR", "SOX9", "MMP7" ),
  macrophage  = c("PTPRC", "LYZ", "CD14", "C1QA", "CD163", "CSF1R", "TREM2", "CD74", "HLA-DRB1"),
  CD4         = c("PTPRC", "CD3G", "CD4"),
  CD8         = c("PTPRC", "CD3G", "CD8A", "CD8B"),
  B           = c("PTPRC","CD79A","CD79B","CD19","MS4A1","IGKC","IGHA1","IGHM","JCHAIN","IGLL5"),  # CD20 = MS4A1
  EC          = c("PECAM1", "VWF", "CDH5", "CLDN5", "SELE", "PLVAP", "VEGFB", "CLDN14"),
  fibro       = c("ACTA2", "VIM", "DES", "COL1A1", "PDGFRA", "FSCN2")
)

# Extract unique marker genes
key_genes <- unique(unlist(panc_markers))

# 🔬 DotPlot of marker expression
DotPlot(nano.obj,
        features = key_genes,
        group.by = "seurat_clusters") +
  # scale_color_viridis_c(option = "D") +  # Options: A, B, C, D, E
  # scale_color_distiller(palette = "RdBu", direction = 1) +
  # scale_color_gradient(low = "#f0f0f0", high = "#084594") +
  scale_color_distiller(palette = "YlGnBu", direction = 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Pancreatic Marker Expression per cell cluster")


# By observing the DotPlot, clusters 15 and 16 are actually sharing the expression of INS, GCG, PPY and SST.
# Thus, see possibility to split and subcluster both clusters and reintegrate them to the rest.


# 🚨 Subclustering: isolate clusters 15 and 16
Idents(nano.obj) <- "seurat_clusters"
sub <- subset(nano.obj, idents = c("15", "16"))

# Re-run PCA + UMAP + Clustering on this subset
sub <- RunPCA(sub, features = VariableFeatures(sub), verbose = FALSE)
sub <- RunUMAP(sub, dims = 1:15, reduction = "pca")
sub <- FindNeighbors(sub, dims = 1:15)
sub <- FindClusters(sub, resolution = 0.5)

# Check subcluster assignments
table(Idents(sub))

# 🧬 Reintegrate refined identities
# Assign new labels like 15.1, 15.2, 16.1, etc.
original_ids <- sub$seurat_clusters
new_sub_ids <- Idents(sub)

# Combine original and subcluster identities
combined_ids <- paste0(as.character(original_ids), ".", as.character(new_sub_ids))

# Set them as identities for the sub object
Idents(sub) <- combined_ids

# Map back into full Seurat object
nano.obj$refined_clusters <- as.character(Idents(nano.obj))  # Start with original
nano.obj$refined_clusters[Cells(sub)] <- as.character(Idents(sub))  # Overwrite subclustered cells

# 🔁 UMAP with refined clusters
p1 <- DimPlot(nano.obj, 
              group.by = "refined_clusters", 
              raster = FALSE, 
              label = TRUE, 
              pt.size = 1.4,
              label.size = 6) + 
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 15)  # Fixed: removed misplaced +
  ) + 
  ggtitle("UMAP: Refined Cell Clustering of Human Pancreas (with Subclusters 15.x & 16.x)") +
  guides(color = guide_legend(
    title = "Cell Cluster ID",
    override.aes = list(size = 4, alpha = 1),
    ncol = 1
  ))

p1

# 🔬 DotPlot using refined clusters
DotPlot(nano.obj,
        features = key_genes,
        group.by = "refined_clusters") +
  scale_color_distiller(palette = "YlGnBu", direction = 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Marker Expression After Refinement")


# ---------------- Azimuth cell annotation --------------------

panc_markers_azimuth <- list(
  beta               = c("INS", "IAPP", "PDX1", "SLC30A8", "DLK1", "INS-IGF2", "G6PC2", "HADH", "ADCYAP1", "GSN", "NPTX2", "C12orf75", "CHGA"),
  alpha              = c("GCG", "TTR", "PPP1R1A", "CRYBA2", "TM4SF4", "MAFB", "GC", "GPX3", "PCSK2", "PEMT"),
  gamma              = c("PPY", "AQP3", "MEIS2", "ID2", "GPC5-AS1", "CARTPT", "PRSS23", "ETV1", "PPY2", "TUBB2A"),
  delta              = c("SST", "RBP4", "SERPINA1", "RGS2", "PCSK1", "SEC11C", "HHEX", "LEPR", "MDK", "LY6H"),
  epsilon            = c("BHMT", "VSTM2L", "PHGR1", "TM4SF5", "ANXA13", "ASGR1", "DEFB1", "GHRL", "COL22A1", "OLFML3", "PYY"),
  acinar             = c("REG1A", "PRSS1", "CTRB2", "CTRB1", "REG1B", "CELA3A", "PRSS2", "REG3A", "CPA1", "CPA2", "MUC6", "CLPS"),
  ductal             = c("SPP1", "MMP7", "IGFBP7", "KRT7", "ANXA4", "SERPINA1", "LCN2", "CFTR", "KRT19", "SERPING1", "KRT8", "KRT7"),
  quiescent_stellate = c("RGS5", "C11orf96", "FABP4", "CSRP2", "IL24", "ADIRF", "NDUFA4L2", "GPX3", "IGFBP4", "ESAM"),
  activated_stellate = c("COL1A1", "COL1A2", "COL6A3", "COL3A1", "TIMP3", "TIMP1", "CTHRC1", "SFRP2", "BGN", "LUM"),
  # schwann            = c("NGFR", "CDH19", "UCN2", "SOX10", "S100A1", "PLP1", "TSPAN11", "WNT16", "SOX2", "TFAP2A"),
  macrophage         = c("LYZ", "CD14", "C1QA", "CD163", "CSF1R", "TREM2", "CD74", "HLA-DRB1", "ACP5", "APOE", "FCER1G", "C1QC", "C1QB", "SRGN"),
  # B                  = c("CD79A","CD79B","CD19","MS4A1","IGKC","IGHA1","IGHM","JCHAIN","IGLL5"), 
  endothelial_cells  = c("PECAM1", "VWF", "CDH5", "CLDN5", "SELE", "PLVAP", "RGCC", "ENG", "ESM1", "SERPINE1", "STC1", "MMP1", "GNG11")
  # fibro              = c("ACTA2", "VIM", "DES", "COL1A1", "PDGFRA", "FSCN2"),
  # cycling            = c("UBE2C", "TOP2A", "CDK1", "BIRC5", "PBK", "CDKN3", "MKI67", "CDC20", "CCNB2", "CDCA3")
)

# Extract unique marker genes
key_genes_azimuth <- unique(unlist(panc_markers_azimuth))

# 🔬 DotPlot of marker expression
DotPlot(nano.obj,
        features = key_genes_azimuth,
        group.by = "refined_clusters") +
  # scale_color_viridis_c(option = "D") +  # Options: A, B, C, D, E
  # scale_color_distiller(palette = "RdBu", direction = 1) +
  scale_color_gradient(low = "#f0f0f0", high = "#084594") +
  # scale_color_distiller(palette = "YlGnBu", direction = 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  ggtitle("Pancreatic Marker Expression per cell cluster")


################################################
##### Flextable: Cluster IDs - Amounts - Freq
################################################

library(patchwork)
library(dplyr)
library(flextable)
library(officer)
library(gridExtra)
library(grid)
library(cowplot)
library(ggplot2)

# 🧮 1. Compute refined cluster frequencies
cluster_summary <- as.data.frame(table(nano.obj$refined_clusters))
colnames(cluster_summary) <- c("Cluster identifier", "Amount of cells")

# 🔽 Sort in descending order
cluster_summary <- cluster_summary %>%
  arrange(desc(`Amount of cells`))

# ➕ 2. Add percentage column
cluster_summary$`% from total` <- round(100 * cluster_summary$`Amount of cells` / sum(cluster_summary$`Amount of cells`), 2)

# ✅ 3. Add total row safely
total_row <- data.frame(
  `Cluster identifier` = "Total",
  `Amount of cells` = sum(cluster_summary$`Amount of cells`),
  `% from total` = 100.00,
  check.names = FALSE  # Prevents column name parsing
)

# 🧼 4. Bind rows explicitly with aligned column names
cluster_summary <- bind_rows(cluster_summary, total_row)

# 🖼️ 5. Format with flextable
ft_summary <- flextable(cluster_summary) %>%
  theme_vanilla() %>%
  add_header_lines("Amount and frequencies of Cell Clusters from human Pancreas*") %>%
  fontsize(i = 1, size = 26, part = "header") %>%
  add_footer_lines(values = c(
    "* Source: https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-human-pancreas-ffpe-dataset/", 
    "* Platform:  CosMx™ Spatial Molecular Imager (SMI).", 
    "* Preservation method: Formalin-Fixed Paraffin-Embedded (FFPE)"
    )) %>% 
  # Alignment
  align(align = "center", part = "header") %>%     # Center header (including title line)
  align(align = "center", part = "body") %>%       # Center all body cells
  align(align = "left", part = "footer") %>%       # Left-align footnote
  # Total row styling
  bold(~ `Cluster identifier` == "Total") %>%
  bg(~ `Cluster identifier` == "Total", bg = "#2D2D2D", part = "body") %>%  
  color(~ `Cluster identifier` == "Total", color = "white", part = "body") %>%  
  # Borders
  border_inner_h(border = fp_border(color = "#D3D3D3", width = 0.5)) %>%
  border_outer(border = fp_border(color = "#D3D3D3", width = 0.5)) %>%
  # Font
  fontsize(size = 22, part = "all") %>%
  autofit()

# Show table
ft_summary <- flextable::font(ft_summary, fontname = "Arial", part = "all")
ft_summary

# Saving image of flextable
# Convert flextable to grob
# ft_grob <- flextable::gen_grob(ft_summary)

# Set the output path
# output_path <- "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/pancreas_umap_table_clusters.png"

# Generate the composite layout with spacing
# library(grid)
# library(gridExtra)

# Manual PNG device creation
# png(output_path, width = 26, height = 14, units = "in", res = 600)
# grid.newpage()
# pushViewport(viewport(width = 0.95, height = 0.95))
# grid.arrange(p1, ft_grob, ncol = 2, widths = c(2, 1.6), padding = unit(2, "cm"))
# dev.off()

# Alternative:
# Set up PNG graphics device
# png("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/pancreas_umap_flextable.png",
#     width = 3600, height = 1800, res = 300)
# 
# # Draw the plot directly to PNG device
# grid.arrange(
#   p1,
#   ft_grob,
#   nrow = 1,
#   widths = c(2, 1.6)
# )
# 
# # Turn off device
# dev.off()



###################################################################################
##### Tissue visualization of genes: INS - GCG - SST - REG1A - SPP1 - C1QC - PECAM1
###################################################################################

ImageFeaturePlot(nano.obj, fov = "Pancreas", 
                 features = "INS", 
                 # pt.size = 0.5,                 # Increase dot size --> It doesn't work
                 max.cutoff = "q95",             # Adjust contrast
                 alpha = 0.8,                   # Semi-transparency
                 cols = c("white", "red1"),       # Custom color gradient
                 size = 1)      

ImageFeaturePlot(nano.obj, fov = "Pancreas", 
                 features = "GCG", 
                 # pt.size = 0.5,                 # Increase dot size --> It doesn't work
                 max.cutoff = "q95",             # Adjust contrast
                 alpha = 0.8,                    # Semi-transparency
                 cols = c("white", "red1"),       # Custom color gradient
                 size = 1)    

ImageFeaturePlot(nano.obj, fov = "Pancreas", 
                 features = c("INS", "GCG", "SST","REG1A","SPP1", "C1QC", "PECAM1"),
                 # pt.size = 0.5,                 # Increase dot size --> It doesn't work
                 max.cutoff = "q95",             # Adjust contrast
                 # alpha = 0.8,                   # Semi-transparency
                 # cols = c("white", "red1"),       # Custom color gradient
                 size = 0.5)

# co-expression
ImageDimPlot(nano.obj,
             fov = "Pancreas", 
             molecules = "INS",
             nmols = 10000,          # Can increase for more visible signal
             alpha = 0.8,            # Less transparency
             size = 1.5) +           # Increase point size
  NoLegend()

p_genes <- ImageDimPlot(
  nano.obj, 
  fov = "Pancreas", 
  molecules = c("INS", "GCG", "GHRL", "PPY"),
  group.by = "refined_clusters",
  # alpha = 0.3, 
  nmols = 10000,
  size = 0.1)

p_genes



# Recreate the plot
library(patchwork)

p_genes_list <- ImageDimPlot(
  nano.obj,
  fov = "Pancreas",
  molecules = c("INS", "GCG", "GHRL", "PPY"),
  group.by = "refined_clusters",
  nmols = 10000,
  size = 0.1,
  combine = FALSE  # 👈 This is the key
)


# ggsave(
#   filename = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/spatial_pancreas_genes_legend.png",
#   plot = p_genes,
#   width = 7,
#   height = 6,
#   dpi = 600
# )

#########################
##### Cell annotation
#########################

# Assign cell types based on refined clusters
nano.obj$cell_type <- case_when(
  nano.obj$refined_clusters %in% c("4.4", "0.0") ~ "Beta",
  nano.obj$refined_clusters %in% c("2.2")        ~ "Delta",
  nano.obj$refined_clusters %in% c("3.3")        ~ "Alpha",
  nano.obj$refined_clusters %in% c("1.1")        ~ "Gamma/B/A",
  nano.obj$refined_clusters %in% c("9", "8", "7", "6", "4", "3", "12", "11", "0") ~ "Acinar",
  nano.obj$refined_clusters %in% c("5", "17", "13", "1") ~ "Ductal",
  nano.obj$refined_clusters %in% c("14", "18")    ~ "Macrophages",
  nano.obj$refined_clusters %in% c("2")           ~ "Stellate",
  nano.obj$refined_clusters %in% c("10")          ~ "Endothelial",
  TRUE ~ "Unassigned"
)

# Check annotation counts
view(nano.obj@meta.data)
table(nano.obj$cell_type)
colnames(nano.obj@meta.data) # the new column "cell_type" with the annotated cells
# [1] "orig.ident"   "nCount_Nanostring"   "nFeature_Nanostring" "percent.mt"      "percent.rb"         
# [6] "nCount_SCT"    "nFeature_SCT"        "SCT_snn_res.1.5"     "seurat_clusters"     "SCT_snn_res.1.2"    
# [11] "SCT_snn_res.1.6"     "SCT_snn_res.1.7"     "SCT_snn_res.1.8"     "SCT_snn_res.2"       "SCT_snn_res.1"  
# [16] "refined_clusters"    "cell_type"    
unique(nano.obj@meta.data$cell_type)
# [1] "Stellate"    "Acinar"      "Ductal"      "Macrophages" "Endothelial" "Alpha"       "Beta"        "Gamma/B/A" 
# [9] "Delta"   

# UMAP
p_celltype <- DimPlot(
  nano.obj, 
  group.by = "cell_type", 
  label = TRUE, 
  repel = TRUE,
  pt.size = 1.5,
  label.size = 6,
  alpha = 0.6,
  label.box = TRUE,
  raster = FALSE
) + 
  ggtitle("UMAP: Annotated Cell Types in Human Pancreas (Nanostring - CosMx)") +
  NoAxes() +
  guides(color = guide_legend(
    title = "Cell Type",
    override.aes = list(size = 4, alpha = 1),
    ncol = 1
  )) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.margin = margin(1, 1, 1, 1, "cm"),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 15),
    legend.key.size = unit(1, "cm"),
    legend.spacing.y = unit(0.4, "cm")
  )

p_celltype

# ggsave(
#   filename = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/pancreas_umap_celltypes.png",
#   plot = p_celltype,
#   width = 13,
#   height = 6,
#   dpi = 600
# )


################################################
##### Flextable: Cell Clusters - Amounts - Freq
################################################

# library(patchwork)
library(dplyr)
library(flextable)
library(officer)
library(gridExtra)
library(grid)
# library(cowplot)
library(ggplot2)

# 🧮 1. Compute cell type frequencies
cluster_summary1 <- as.data.frame(table(nano.obj$cell_type))
colnames(cluster_summary1) <- c("Type of pancreatic cells", "Amount of cells")

# 🔽 2. Sort in descending order
cluster_summary1 <- cluster_summary1 %>%
  arrange(desc(`Amount of cells`))

# ➕ 3. Add percentage column
cluster_summary1$`Frequency from total (%)` <- round(
  100 * cluster_summary1$`Amount of cells` / sum(cluster_summary1$`Amount of cells`), 2
)

# ✅ 4. Add total row
total_row <- data.frame(
  `Type of pancreatic cells` = "Total",
  `Amount of cells` = sum(cluster_summary1$`Amount of cells`),
  `Frequency from total (%)` = 100.00,
  check.names = FALSE
)

cluster_summary1 <- bind_rows(cluster_summary1, total_row)

# 🖼️ 5. Format with flextable
ft_summary1 <- flextable(cluster_summary1) %>%
  theme_vanilla() %>%
  add_header_lines("Content and types of pancreatic cells from Human*") %>%
  fontsize(i = 1, size = 26, part = "header") %>%
  add_footer_lines(values = c(
    "* Source: https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-human-pancreas-ffpe-dataset/", 
    "* Platform: CosMx™ Spatial Molecular Imager (SMI).", 
    "* Preservation method: Formalin-Fixed Paraffin-Embedded (FFPE)"
  )) %>%
  align(align = "center", part = "header") %>%
  align(align = "center", part = "body") %>%
  align(align = "left", part = "footer") %>%
  bold(~ `Type of pancreatic cells` == "Total", part = "body") %>%
  bg(~ `Type of pancreatic cells` == "Total", bg = "#2D2D2D", part = "body") %>%
  color(~ `Type of pancreatic cells` == "Total", color = "white", part = "body") %>%
  border_inner_h(border = fp_border(color = "#D3D3D3", width = 0.5)) %>%
  border_outer(border = fp_border(color = "#D3D3D3", width = 0.5)) %>%
  fontsize(size = 16, part = "all") %>%
  font(fontname = "Arial", part = "all") %>%
  fontsize(size = 10, part = "footer") %>%
  autofit()

ft_summary1


# Convert flextable to a grid grob
# ft_grob <- flextable::gen_grob(ft_summary1)

# Set the output path
# output_path_table <- "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/pancreas_celltypes_table_alone1.png"

# Create a white background plot using grid
# png(output_path_table, 
#     width = 1300, 
#     height = 800,
#     res = 600,
#     bg = "white")  # Adjust dimensions as needed
# grid::grid.newpage()
# grid::grid.rect(gp = grid::gpar(fill = "white", col = "white"))  # White background
# grid::grid.draw(ft_grob)
# dev.off()


#########################################
##### Tissue visualization of cell types
#########################################

# Cell types
ImageDimPlot(
  nano.obj,
  fov = "Pancreas",
  group.by = "cell_type",   # 🧠 correct metadata field
  # cols = "glasbey",
  cells = WhichCells(nano.obj, expression = cell_type %in% 
                       c("Beta", "Alpha", "Macrophages", "Acinar", "Ductal")),
  nmols = 5000,
  size = 1.1
)

# [1] "Stellate"    "Acinar"      "Ductal"      "Macrophages" "Endothelial" "Alpha"       "Beta"        "Gamma/B/A" 
# [9] "Delta"


# -------------------- Changing the background to white ------------------------


# Full plot with both panels and legend
p_spatial <- ImageDimPlot(
  nano.obj,
  fov = "Pancreas",
  group.by = "cell_type",
  cells = WhichCells(nano.obj, expression = cell_type %in% 
                       c("Beta", "Alpha", "Macrophages", "Acinar", "Ductal")),
  nmols = 10000,
  size = 1.0
) & 
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  )

p_spatial

# ggsave(
#   filename = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/spatial_pancreas_celltypes_legend.png",
#   plot = p_spatial,
#   width = 7,
#   height = 6,
#   dpi = 600
# )



# ------------------------ Cell types - Zoom in -----------------------------

library(future)

options(future.globals.maxSize = 10 * 1024^3)  # 10 GB


### 1. Initial full-tissue overview
ImageDimPlot(nano.obj, fov = "Pancreas", 
             group.by = "cell_type",
             nmols = 10000,
             size = 1.0,
             axes = TRUE) # cols = "glasbey")

### 2. Crop region (adjust `x` and `y` ranges to zoom into desired area)
pancreas.crop <- Crop(nano.obj[["Pancreas"]],
                      x = c(31000, 35000),   # <-- adjust as needed
                      y = c(61000, 64000))   # <-- adjust as needed

### 3. Add new zoomed FOV to object
nano.obj[["zoom_pancreas"]] <- pancreas.crop

### 4. Define segmentation boundaries if not auto-detected
DefaultBoundary(nano.obj[["zoom_pancreas"]]) <- "segmentation"

### 5. Visualize zoomed region: colored by cell type (or cluster)
p_spatial_zoom <- ImageDimPlot(
  nano.obj,
  fov = "zoom_pancreas",
  # cols = "glasbey",            # use your preferred palette
  group.by = "cell_type",      # or "refined_clusters"
  # cols = "glasbey",
  alpha = 0.8,
  coord.fixed = FALSE
)

p_spatial_zoom

# ggsave(
#   filename = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/spatial_pancreas_celltypes_zoom_legend.png",
#   plot = p_spatial_zoom,
#   width = 7,
#   height = 6,
#   dpi = 600
# )

### 5.1 Visualize zoomed region: colored by cell type (or cluster): Changing palette
p_spatial_zoom_glasbey <- ImageDimPlot(
  nano.obj,
  fov = "zoom_pancreas",
  cols = "glasbey",            # use your preferred palette
  group.by = "cell_type",      # or "refined_clusters"
  # cols = "glasbey",
  alpha = 0.8,
  coord.fixed = FALSE
)

p_spatial_zoom_glasbey

# ggsave(
#   filename = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/spatial_pancreas_celltypes_zoom_legend_gl.png",
#   plot = p_spatial_zoom_glasbey,
#   width = 7,
#   height = 6,
#   dpi = 600
# )

### 6. Add molecule-level visualization for markers of interest

mol_zoom1 <- ImageFeaturePlot(             # It works!!
  nano.obj,
  features = c("INS", "GCG", "GHRL", "PPY"),
  fov = "zoom_pancreas",
  alpha = 1.0,
  max.cutoff = "q95",         # Contrast control
  size = 1.2,                 # Marker size
  cols = c("white", "red1")   # Color gradient
)

# ggsave(
#   filename = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/spatial_pancreas_zoom_molecules1_legend.png",
#   plot = mol_zoom1,
#   width = 7,
#   height = 6,
#   dpi = 600
# )


ImageDimPlot(
  nano.obj,
  fov = "Pancreas",
  group.by = "cell_type",
  molecules = c("INS", "GCG", "GHRL", "PPY"),
  mols.size = 1.0,                             # 🔸 Slightly larger molecule dots
  nmols = 15000,                               # 🔸 More molecules (adjust for crowding)
  size = 2,                                    # 🔸 Increase cell boundary thickness
  border.color = "black",                       # 🔸 Makes cells stand out
  alpha = 0.7
) + 
  coord_cartesian(
    xlim = c(31000, 35000),
    ylim = c(61000, 64000)
  ) + 
  ggtitle("Zoomed-In View: Cell Types and Key Hormones") +
  theme_minimal(base_size = 14)


molzoom2 <- ImageDimPlot(
  nano.obj,
  fov = "Pancreas",
  group.by = "cell_type",
  molecules = c("INS", "GCG", "GHRL", "PPY", "SST", "HLA-DRB1", "PECAM1", "KRT8"),
  mols.size = 1.0,            
  nmols = 15000,              
  size = 0.6,                   
  border.color = "white",    
  alpha = 0.4
) + 
  coord_cartesian(
    xlim = c(31000, 35000),
    ylim = c(61000, 64000)
  ) +
  ggtitle("Zoomed-In Pancreas: Key Hormones and other molecules") +
  theme(
    panel.background = element_rect(fill = "black", color = NA),
    plot.background  = element_rect(fill = "black", color = NA),
    legend.background = element_rect(fill = "black"),
    legend.key = element_rect(fill = "black"),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    plot.title = element_text(color = "white", size = 14, face = "bold"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white")
  )

molzoom2



# Rebuild plot with combined legend
molzoom2 <- ImageDimPlot(
  nano.obj,
  fov = "Pancreas",
  group.by = "cell_type",
  molecules = c("INS", "GCG", "GHRL", "PPY", "SST", "HLA-DRB1", "PECAM1", "KRT8"),
  mols.size = 1.0,
  nmols = 15000,
  size = 0.6,
  border.color = "white",
  alpha = 0.4
) +
  coord_cartesian(
    xlim = c(31000, 35000),
    ylim = c(61000, 64000)
  ) +
  ggtitle("Zoomed-In Pancreas: Key Hormones and cell markers") +
  theme(
    panel.background = element_rect(fill = "black", color = NA),
    plot.background  = element_rect(fill = "black", color = NA),
    legend.background = element_rect(fill = "black"),
    legend.key = element_rect(fill = "black"),
    legend.text = element_text(color = "white"),
    legend.title = element_text(color = "white"),
    plot.title = element_text(color = "white", size = 14, face = "bold"),
    axis.text = element_text(color = "white"),
    axis.title = element_text(color = "white")
  )

# Try suppressing the cell_type legend manually by name
molzoom2_clean <- molzoom2 +
  guides(
    fill = "none",     # remove fill legend (often for cell_type in ImageDimPlot)
    color = guide_legend(override.aes = list(size = 2))  # keep molecules
  )

molzoom2_clean


# ggsave(
#   filename = "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/spatial_pancreas_zoom_molecules3_legend.png",
#   plot = molzoom2_clean,
#   width = 7,
#   height = 6,
#   dpi = 600
# )


# --------

# nano.obj
# An object of class Seurat 
# 37992 features across 48944 samples within 2 assays 
# Active assay: SCT (18996 features, 5000 variable features)
#  3 layers present: counts, data, scale.data
#  1 other assay present: Nanostring
#  2 dimensional reductions calculated: pca, umap
#  2 spatial fields of view present: Pancreas zoom_pancreas

# Save
# saveRDS(nano.obj, "/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_annotated.rds")

library(Seurat)
# Calling nano.obj
# nano.obj <- readRDS("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/nano_annotated.rds")

# nano.obj
# An object of class Seurat 
# 37992 features across 48944 samples within 2 assays 
# Active assay: SCT (18996 features, 5000 variable features)
#  3 layers present: counts, data, scale.data
#  1 other assay present: Nanostring
#  2 dimensional reductions calculated: pca, umap
#  2 spatial fields of view present: Pancreas zoom_pancreas

```

## *Conversion into H5AD format*

```{r}

view(nano.obj@meta.data)
colnames(nano.obj@meta.data)
# [1] "orig.ident"     "nCount_Nanostring"   "nFeature_Nanostring" "percent.mt"     "percent.rb"         
# [6] "nCount_SCT"     "nFeature_SCT"        "SCT_snn_res.1.5"     "seurat_clusters"     "SCT_snn_res.1.2"    
# [11] "SCT_snn_res.1.6"     "SCT_snn_res.1.7"     "SCT_snn_res.1.8"     "SCT_snn_res.2"       "SCT_snn_res.1"      
# [16] "refined_clusters"    "cell_type"  

# Set assay
DefaultAssay(nano.obj) <- "SCT"

# Extract centroids (x = imagecol, y = imagerow)
# Step 1: Get spatial coordinates
coords <- GetTissueCoordinates(nano.obj, fov = "Pancreas")
head(coords)
#          x        y cell
# 1 52580.04 41889.41 1_51
# 2 52711.04 41908.41 2_51
# 3 52802.04 41899.41 3_51
# 4 52908.04 41880.41 4_51
# 5 53011.04 41885.41 5_51
# 6 53165.04 41904.41 6_51

# Step 2: Fix rownames to match cell barcodes (required for matching to Seurat cells)
rownames(coords) <- coords$cell

# Add to meta.data (important for .h5ad)
# Step 3: Now safely assign x/y to meta.data
nano.obj$x <- coords[Cells(nano.obj), "x"]
nano.obj$y <- coords[Cells(nano.obj), "y"]

# Confirmation that coordinates were add it to meta.data
colnames(nano.obj@meta.data)
# [1] "orig.ident"          "nCount_Nanostring"   "nFeature_Nanostring" "percent.mt"         
#  [5] "percent.rb"          "nCount_SCT"          "nFeature_SCT"        "SCT_snn_res.1.5"    
#  [9] "seurat_clusters"     "SCT_snn_res.1.2"     "SCT_snn_res.1.6"     "SCT_snn_res.1.7"    
# [13] "SCT_snn_res.1.8"     "SCT_snn_res.2"       "SCT_snn_res.1"       "refined_clusters"   
# [17] "cell_type"           "x"                   "y"    

view(nano.obj@meta.data)

head((nano.obj@meta.data), 3)

head(nano.obj@meta.data[, c("x", "y")], 3)
#    x        y
# 1_51 52580.04 41889.41
# 2_51 52711.04 41908.41
# 3_51 52802.04 41899.41


# -------------------------- H5AD conversion --------------------------

# Load necessary packages
library(Seurat)
library(SingleCellExperiment)
library(zellkonverter)
# library(reticulate)

# Convert Seurat to SingleCellExperiment (spatial coords now inside meta.data)
sce <- as.SingleCellExperiment(nano.obj)

# Ensure output folder exists
# dir.create("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2", showWarnings = FALSE, recursive = TRUE)

# Save to H5AD format
output_path <- file.path("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/nano_annotated.h5ad")
writeH5AD(sce, file = output_path)

message("✅ H5AD successfully written with spatial coordinates to: ", output_path)

# -------------------------- H5AD conversion: lighter version --------------------------

# Reduce object size before conversion
nano.obj_light <- DietSeurat(nano.obj, 
                       counts = TRUE, 
                       data = TRUE, 
                       scale.data = FALSE, 
                       assays = "SCT")

sce <- as.SingleCellExperiment(nano.obj_light)
output_path <- file.path("/home/ubuntu/proj6_nanoString_CosMx_Spatial_Pancreas/proj6_RData_EC2/nano_light_annotated.h5ad")
writeH5AD(sce, file = output_path)

message("✅ H5AD successfully written with spatial coordinates to: ", output_path)

# Because the layer `scale.data` was not included in "nano.obj_light",
# the object then only keeps the `SCT` assay with `counts` and `data` layers.
# The "nano_annotated.h5ad" is 7.5GB while the "nano_light_annotated.h5ad" only 535MB.
# This is completely expected and demonstrates a key point about spatial data storage. 

# Key Insights:
# 
#     The 7.5GB file contains:
# 
#         Full gene expression matrices (counts + data)
# 
#         Pearson residuals (scale.data)
# 
#         All metadata columns
# 
#         Spatial polygon coordinates
# 
#         Image data references
# 
#     The 535MB light version contains:
# 
#         Only raw counts and normalized data
# 
#         Basic metadata (no polygons)
# 
#         Cluster assignments
# 
#         Spatial coordinates as single points (not polygons)

# Alternative DietSeurat()
# Best practice for export
# nano.obj_export <- DietSeurat(
#   nano.obj,
#   assays = "SCT",
#   layers = c("counts", "data"),  # Exclude scale.data
#   dimreducs = c("pca", "umap"),  # Keep reductions
#   graphs = NULL                  # Exclude neighbor graphs
# )
#
#For Spatial Analysis in Python:

    # Use the light version (535MB) for most workflows
    # 
    # Only keep the full version if you need:
    # 
    #     Detailed polygon boundaries
    # 
    #     Scale.data for exact Pearson residuals
    # 
    #     Image data embedded in the file

# Final Size Comparison:
# Component	          Full Version   	Light Version
# Gene Expression	      ✓ ✓ ✓	            ✓ ✓
# Scale Data	           ✓              	✗
# Spatial Polygons	     ✓	              ✗
# Image Data	           ✓              	✗
# Reductions	           ✗	              ✗
# File Size	           7.5 GB	          535 MB
# Python Loading	    Minutes	         Seconds

# Reductions: refers to dimentionality reductions like PCA, UMAP, etc.SPCA (Spatial PCA), tSNE

# Component	            Full Version	              Light Version	              Size Impact
# scale.data	              ✓ (dense matrix)	          ✗	                        ~7 GB
# Spatial Polygons	        ✓ (detailed)	              ✗ (only centroids)	      ~0.5 GB
# Image Data	              ✓ (references)            	✗	                        ~0.2 GB
# Other Assays	            ✓ (Nanostring)	            ✗	                        ~1 GB
# Metadata	                ✓ (full)	                  ✓ (reduced)	              Minimal
# Reductions	              ✗	                          ✗	                          -



```

```{r}

```
